### 2025-06-22 manuscript/main.tex
requirements and alterantive technologies

\usepackage[backend=biber, style=numeric]{biblatex}
\addbibresource{bibliography/theory.bib}

@@

\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} ?NAME? for \textit{MiniProb}. Language extensions are a big part of
VS Code’s extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts, validating the code in the process. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.

@@

\chapter{Requirements \& Technology Survey}
\section{Functional Requirements}
\section{Non-functional Requirements}
\section{Alternative Technologies}
\section{Justification for Langium \& VS Code}

\chapter{Architectural Design}
\section{High-Level Architecture Diagram}
\section{Module Decomposition}
\section{Data Flow and Control Flow}
\section{Language Server Protocol}

@@

\chapter{Implementation Details}
\label{sec:langium-grammar}
\section{Langium Grammar Definition for \textit{Miniprob}}
\section{Semantic Checks and Type System}
\section{VS Code Extension Points}

@@

\chapter{Discussion}
\section{Meeting the Requirements}

>>>>

\usepackage[backend=biber, style=numeric]{biblatex}
\addbibresource{bibliography/theory.bib}
\addbibresource{bibliography/implementation.bib}

@@

\label{sec:scope}
\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} ?NAME? for \textit{MiniProb}. Language extensions are a big part of
VS Code’s extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts, validating the code in the process. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.

@@

\chapter{Requirements and Technological Context}

As discussed in in the introduction \ref{sec:scope}, the main aim is the development of a VS Code Language extension that operates on par with
well-established tooling for other programming languages. The following requirements define the expected capabilities and characteristics the extension must demonstrate.

\begin{multicols}{2}
  \subsection*{Functional Requirements}
  \begin{enumerate}
    \item The extension must provide syntax highlighting for the target language, differentiating keywords, operators, literals, and identifiers.
    \item The extension must provide basic code completion suggestions based on language keywords and the abstract syntax tree (AST) generated by Langium.
    \item The extension must be able to validate data types within the language, providing warnings or errors when type mismatches occur.
    \item The extension must perform semantic analyses to detect invalid constructs, undefined variables, and other semantic errors.
    \item The extension must display syntax and semantic error messages in the editor in real-time as the user edits the code.
    \item The extension must utilize the Langium framework for language specification, parser generation, and language service implementation.
  \end{enumerate}

  \columnbreak

  \subsection*{Non-Functional Requirements}
  \begin{enumerate}
    \item The extension must respond to user edits with syntax and semantic feedback within 100-250ms for files of average size.
    \item The extension must provide a user-friendly experience that is consistent with other popular VS Code extensions.
    \item The extension must be structured and commented in a way that allows future developers to understand and extend its functionality easily.
    \item The extension must be compatible with the latest version of Visual Studio Code at the time of its release and be developed using Langium, with consideration for future maintenance and version support.
    \item The extension must be able to handle files of immense size (around 10000 lines) of code within the bounds of 1 second.
    \item The extension must be delivered with clear and comprehensive documentation, including installation, usage, and extension of its features.
  \end{enumerate}
\end{multicols}

\section{Alternative Technologies}

Although this extension is implemented using \textbf{Langium}, which itself builds upon the standardized \textbf{Language Server Protocol (LSP)}~\cite{LSP},
there are several other technologies and approaches available for creating Visual Studio Code language extensions. At its core, the LSP allows tooling for programming
languages to be developed in a modular, editor-independent manner, making it possible to implement a language server in any language that can communicate via JSON-RPC.

One common approach is to implement a language server using general-purpose programming languages (e.g., TypeScript, Java, or Rust) combined with parser generation libraries.
Technologies such as \textbf{Jison}~\cite{Jison}, \textbf{Nearley}~\cite{Nearley}, and \textbf{PEG.js}~\cite{PEGjs} offer a range of options for specifying a language's grammar
and creating a parser. These libraries vary in design and capabilities and may need extra components, such as dedicated lexers and scoping to enable fully-fletched tooling.

In contrast, more comprehensive technologies such as \textbf{ANTLR}~\cite{ANTLR} and \textbf{Langium} (built upon the \textbf{Chevrotain} parser~\cite{Chevrotain}) go beyond
basic parsing. ANTLR provides a mature and highly optimized lexer and parser generation framework that is widely used across many programming languages, while Langium offers
an end-to-end solution for language engineering - including built-in support for scoping, indexing, and validation - making it especially suited for creating fully
featured language servers.

For basic syntax highlighting and lexical analysis, \textbf{TextMate}-style grammars~\cite{TextMate} can be used, offering a lightweight approach when deep semantic understanding is not required.
Meanwhile, incremental parsing frameworks such as \textbf{Tree-sitter}~\cite{Treesitter} have gained popularity for their ability to provide high-performance syntax parsing and error recovery,
making them ideal for tooling that operates on very large files.

Each approach has its benefits and trade-offs, and the choice depends upon the specific requirements, constraints, and long-term maintainability of the language tooling in question.

\chapter{Langium}
\section{High-Level Architecture Diagram}
\section{Module Decomposition}
\section{Data Flow and Control Flow}
\section{Language Server Protocol}

@@

\chapter{Implementation Details}
\label{sec:langium-grammar}
\section{Langium Grammar Definition for \textit{Miniprob}}
\section{Type System and Semantic Checks}
\section{VS Code Extension Points}
textmate highlighting
\section{Testing \& Validation}

@@

\chapter{Discussion?}
\section{Meeting the Requirements}
